---
title: "FDS Project"
output:
  word_document: default
  html_notebook: default
---

Load data set.
```{r}
setwd("F:/Data Science/Project") # change directory
covid = read.csv("covid.csv", na.strings = "?", stringsAsFactors = T) # read data set

covid[is.na(covid)] = 0 # Replace missing values with 0
```

```{r}
summary(covid)
```

Q1) Are there any geographical clusters of high Covid-19 hospitalizations?
```{r}
library(ggplot2)

# Create a ggplot object using the 'covid' data frame, mapping the 'state' variable to the # x-axis and 'total_adult_patients_hospitalized_confirmed_and_suspected_covid' variable to # the y-axis
ggplot(covid, aes(x = state, y = total_adult_patients_hospitalized_confirmed_and_suspected_covid)) +
  
  # Add a bar geometry to the plot, representing the data as bars
  # 'stat = "identity"' is used when the y values are the actual heights of the bars
  geom_bar(stat = "identity", fill = "skyblue") +
  
  # Set the title, x-axis label, and y-axis label for the plot
  labs(title = "Covid-19 Hospitalizations by State",
       x = "State",
       y = "Total hospitalizations") +
  
  # Rotate for better reading
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
CA has the highest hospitalizations but the adjacent states have very less count.
From my observation of the chart, NC, SC, FL, GA, VA and PA form a good cluster of states with abundant number of hospitalizations.


Q2) Which regions/state might need more resources to deal with the pandemic?
```{r}
library(ggplot2)

# Create a dot plot of staffing shortage by state
ggplot(covid, aes(x = state, y = critical_staffing_shortage_anticipated_within_week_yes)) +
  
  # Add points to the plot, representing the data as dots
  # 'size = 3' sets the size of the dots
  geom_point(size = 3, color = "skyblue") +
  
  # Set the title, x-axis label, and y-axis label for the plot
  labs(title = "States with Anticipated Critical Staffing Shortage within a Week",
    x = "State",
    y = "Number of hospitals anticipating staffing shortage") +
  
    # Rotate for better reading
    theme(axis.text.x = element_text(angle = 45, hjust = 1))



# Create another dot plot of inpatient beds utilization by state
ggplot(covid, aes(x = state, y = inpatient_beds_utilization)) +
  geom_point(size = 3, color = "skyblue") +
  labs(title = "States w.r.t.o inpatient beds utilization",
    x = "State",
    y = "Percent of inpatient beds being utilized") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
CA and MI are the states that needs more staff within the next week.
RI has the highest percentage of inpatient beds utilized, while WY has the least. So, some beds from WY can be shifted to RI as backup.


Q3) Any regional differences in ICU bed utilization?
```{r}
library(ggplot2)

# Create a box plot of adult ICU bed utilization by state
ggplot(covid, aes(x = state, y = adult_icu_bed_covid_utilization)) +
  geom_boxplot() +
  labs(title = "ICU Bed Utilization by State (Adult)", x = "State", y = "ICU Bed Utilization") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Create a box plot of total staffed pediatric ICU bed coverage by state
ggplot(covid, aes(x = state, y = total_staffed_pediatric_icu_beds_coverage)) +
  geom_boxplot() +
  labs(title = "ICU Bed Utilization by State (Pediatric)", x = "State", y = "ICU Bed Utilization") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
AZ and MD are on top of the list when it comes to adult ICU beds utilization.
CA and TX are on top of the list when it comes to pediatric ICU beds utilization.



Q4) Does age have a say in Covid-19 hospitalizations? How different do pediatrics and adults fare?
```{r}
library(dplyr)
library(tidyr)

# Summarize the data by age group
covid_summary <- covid %>%
  summarise(
    age_18_19 = sum(previous_day_admission_adult_covid_confirmed_18.19),
    age_20_29 = sum(previous_day_admission_adult_covid_confirmed_20.29),
    age_30_39 = sum(previous_day_admission_adult_covid_confirmed_30.39),
    age_40_49 = sum(previous_day_admission_adult_covid_confirmed_40.49),
    age_50_59 = sum(previous_day_admission_adult_covid_confirmed_50.59),
    age_60_69 = sum(previous_day_admission_adult_covid_confirmed_60.69),
    age_70_79 = sum(previous_day_admission_adult_covid_confirmed_70.79),
    age_80_plus = sum(previous_day_admission_adult_covid_confirmed_80.),
    age_unknown = sum(previous_day_admission_adult_covid_confirmed_unknown)
  )

# Display the summary
print(covid_summary)

# Find age group with the most and least patients
max_age_group <- colnames(covid_summary)[which.max(covid_summary)]
min_age_group <- colnames(covid_summary)[which.min(covid_summary)]

print(paste("Age group with most patients:", max_age_group))
print(paste("Age group with least patients:", min_age_group))


# Calculate the mean counts in all columns
mean_counts <- covid %>%
  summarise(
    mean_18_19 = mean(previous_day_admission_adult_covid_confirmed_18.19),
    mean_20_29 = mean(previous_day_admission_adult_covid_confirmed_20.29),
    mean_30_39 = mean(previous_day_admission_adult_covid_confirmed_30.39),
    mean_40_49 = mean(previous_day_admission_adult_covid_confirmed_40.49),
    mean_50_59 = mean(previous_day_admission_adult_covid_confirmed_50.59),
    mean_60_69 = mean(previous_day_admission_adult_covid_confirmed_60.69),
    mean_70_79 = mean(previous_day_admission_adult_covid_confirmed_70.79),
    mean_80_plus = mean(previous_day_admission_adult_covid_confirmed_80.),
    mean_unknown = mean(previous_day_admission_adult_covid_confirmed_unknown)
  )

# Display the mean counts
print(mean_counts)


# Create box plots to visualize the distribution of admissions for each group
boxplot(
  covid$previous_day_admission_adult_covid_confirmed,
  covid$previous_day_admission_pediatric_covid_confirmed,
  names = c("Adults", "Pediatrics"),
  main = "COVID Admissions Comparison",
  ylab = "Number of Admissions"
)

# Perform a t-test to compare means between the two groups
t_test_result <- t.test(
  covid$previous_day_admission_adult_covid_confirmed,
  covid$previous_day_admission_pediatric_covid_confirmed
)

# Display the t-test result
print(t_test_result)

```
Also, p-value is less than 0.05 and that means the variable is statistically significant.


Q5) Any Correlation between beds occupied and deaths recorded?
```{r}
# Calculate correlation coefficient
cor(covid$deaths_covid, covid$inpatient_beds_used_covid)

# Scatter plot
pairs(~  deaths_covid + inpatient_beds_used_covid, data = covid)

# Compute correlation matrix
cor_matrix = cor(covid[, c("deaths_covid", "inpatient_beds_used_covid")])

library(corrplot)
corrplot(cor_matrix, method = "color") # Heat map (color coded correlation plot)
```
Strong positive correlation between the above variables, meaning they are closely associated.


Q6) Any correlation between suspected Covid-19 the previous day and the confirmed cases the subsequent day?
```{r}
# Compute correlation coefficient
cor(covid$previous_day_admission_adult_covid_suspected, covid$total_adult_patients_hospitalized_confirmed_covid)

# Plot heat map (number coded)
cor_matrix = cor(covid[, c("previous_day_admission_adult_covid_suspected", "total_adult_patients_hospitalized_confirmed_covid")])
library(corrplot)
corrplot(cor_matrix, method = "number")

# Regression model to understand the significance of variables
model <- lm(total_adult_patients_hospitalized_confirmed_covid ~ previous_day_admission_adult_covid_suspected, data = covid)
summary(model)
```
Both the variables have weak positive correlation.
Also, as p-values are lesser than 0.05, they are statistically significant.


Extra - (Correlation between confirmed covid cases and covid deaths)
```{r}
cor(covid$deaths_covid, covid$hospital_onset_covid)
```
Again, it is moderately positive correlation, meaning the variables are weakly associated.
